Tight battle between republicans and democrats with Kamala in the lead

### 2024-11-05 paper/paper.qmd
Fixed formatting issues
Corrected figure 3,4,13,14 to be in the correct section. Fixed citation.

The dataset was read into R [@citeR], and cleaned using the R package tidyverse [@tidyverse]. Due to Kamal's late entry into the race, some polls do not have any data on her specifically. This is why we set a cut-off date at 2024/07/21, the day Kamala replaced Joe Biden as the Democratic nominee.

@@

@fig_pollster-harris & @fig_pollster_trump show an interesting bump in predictions by George Mass. This may have been an error on their part however, as it is almost immediately rectified.

```{r}
#| label: fig-pollster_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. Each graph shows individual polls conducted in the months leading up to the election. They represent the change in prediciton for Kamala Harris winning the presidency.
#| echo: false
#| warning: false
#| message: false

# Facet by pollster
base_plot_kamala +
  geom_point() +
  geom_smooth(color = "blue") +
  facet_wrap(vars(pollster))
```

```{r}
#| label: fig-pollster_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. Each graph shows individual polls conducted in the months leading up to the election. They represent the change in prediciton for Donald Trump winning the presidency.  
#| echo: false
#| warning: false
#| message: false

@@

### `states`

@@

@fig-state has Harris win 4 t 7 battleground states, but the margins are increadibly close and all it would take is for one of them to be incorrect to reverse the prediction.

@@

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```
## Key Metrics for Model 1

Table 5 presents the key metrics for Model 1, which served as a baseline analysis for polling percentages for Kamala Harris and Donald Trump. The R-squared values are relatively low for both candidates, at 0.07 for Harris and 0.14 for Trump, indicating that this simple model explained only a small proportion of the variance in polling percentages. The AIC values were 4926.19 and 5053.99 for Harris and Trump respectively, with similar RMSE values around 4, suggesting limited predictive power. This baseline model was used to establish a simple trend before moving to a more comprehensive model that accounts for additional variability.
```{r}
#| echo: false
#| eval: true
#| label: Summary Statistics for Donald Trump
#| tbl-cap: "Summary Statistics for Donald Trump"
#| warning: false
library(modelsummary)
library(knitr)
summary_trump_model_2 <- summary(model_date_pollster_trump)
significant_trump_model_2 <- as.data.frame(summary_trump_model_2$coefficients) %>%
  filter(`Pr(>|t|)` < 0.05)

```{r}
#| echo: false
#| eval: true
#| label: Summary Statistics for Kamala Harris
#| tbl-cap: "Summary Statistics for Kamala Harris"
#| warning: false
library(modelsummary)
library(knitr)
summary_harris_model_2 <- summary(model_date_pollster)
significant_harris_model_2 <- as.data.frame(summary_harris_model_2$coefficients) %>%
  filter(`Pr(>|t|)` < 0.05)

@@

Table 6 and Table 7 present the summary statistics for Model 2 for Donald Trump and Kamala Harris respectively. The second model extends the baseline model by including pollster-specific effects, which significantly improves its predictive capabilities.

@@

```{r}
#| echo: false
#| eval: true
#| label: Polling Prediction Outcome
#| tbl-cap: "Polling prediction Outcome"
#| warning: false
# Plot model predictions for Trump
# Combine the datasets for Trump and Harris, keeping their respective predictions
combined_data_with_preds <- bind_rows(
  just_harris_high_quality |> mutate(candidate = "Kamala Harris"),
  just_trump_high_quality |> mutate(candidate = "Donald Trump")
)

@@

  # Add labels and theme
  labs(
    x = "Polling End Date ",
    y = "Polling Percentage",
    title = "Polling Predictions for Kamala Harris and Donald Trump",
    color = "Candidate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Plotting the smoothed predictions for both Trump and Harris
ggplot(combined_data_with_preds, aes(x = end_date, y = pct)) +
  # Actual data points with transparency
  geom_point(aes(color = candidate), alpha = 0.3) +

@@

  # Add labels and theme
  labs(
    x = "Polling End Date",
    y = "Polling Percentage",
    title = "Smoothed Polling Predictions for Kamala Harris and Donald Trump",
    color = "Candidate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Figure 1 and Figure 2 depict the predicted polling percentages for both candidates over time based on Model 2. The solid lines represent the smoothed predictions for Kamala Harris (blue) and Donald Trump (red). The visualizations reveal that both candidates maintain relatively stable trends, with some fluctuation, particularly influenced by pollsters and specific dates.

The smoothed polling predictions in Figure 2 indicate that Kamala Harris tends to have a slightly higher polling percentage compared to Donald Trump, with less fluctuation over the observed period. This suggests that her polling results are more consistent across different pollsters and dates.

@@

The smoothed polling predictions (Figure 2) clearly showed that Harris is ahead of Trump throughout the analyzed period. This consistent lead, combined with higher polling percentages reported by major polling organizations such as MassINC Polling Group and University of Maryland/Washington Post, indicates Harris’s current advantage heading toward the election.

@@

# Model details {#sec-model-details}

## Posterior predictive checks

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

#plot(first_model, "trace")

#plot(first_model, "rhat")
```

>>>>

The dataset was read into R [@citeR], and cleaned using the R package tidyverse [@tidyverseReference]. Due to Kamal's late entry into the race, some polls do not have any data on her specifically. This is why we set a cut-off date at 2024/07/21, the day Kamala replaced Joe Biden as the Democratic nominee.

@@

@fig-pollster_harris & @fig-pollster_trump show an interesting bump in predictions by George Mass. This may have been an error on their part however, as it is almost immediately rectified.
```{r}
#| label: fig-pollster_harris
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. Each graph shows individual polls conducted in the months leading up to the election. They represent the change in prediciton for Kamala Harris winning the presidency.
#| echo: false
#| warning: false
#| message: false

# Facet by pollster
base_plot_kamala +
  geom_point() +
  geom_smooth(color = "blue") +
  facet_wrap(vars(pollster))
```
```{r}
#| label: fig-pollster_trump
#| fig-cap: As election day draws close, pollsters are wrapping up their polls and posting their predictions on who will win. Each graph shows individual polls conducted in the months leading up to the election. They represent the change in prediciton for Donald Trump winning the presidency.  
#| echo: false
#| warning: false
#| message: false

@@

{{< pagebreak >}}
### `states`

@@

@fig-state has Harris win 4 t0 7 battleground states, but the margins are increadibly close and all it would take is for one of them to be incorrect to reverse the prediction.

@@

# Results
## Key Metrics for Model 1

@tbl-keymetrics-model1 presents the key metrics for Model 1, which served as a baseline analysis for polling percentages for Kamala Harris and Donald Trump. The R-squared values are relatively low for both candidates, at 0.07 for Harris and 0.14 for Trump, indicating that this simple model explained only a small proportion of the variance in polling percentages. The AIC values were 4926.19 and 5053.99 for Harris and Trump respectively, with similar RMSE values around 4, suggesting limited predictive power. This baseline model was used to establish a simple trend before moving to a more comprehensive model that accounts for additional variability.

```{r}
#| label: tbl-keymetrics-model1
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

@@

@tbl-summarystatistics-trump and @tbl-summarystatistics-kamala present the summary statistics for Model 2 for Donald Trump and Kamala Harris respectively. The second model extends the baseline model by including pollster-specific effects, which significantly improves its predictive capabilities.

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-summarystatistics-trump
#| warning: false
library(modelsummary)
library(knitr)
summary_trump_model_2 <- summary(model_date_pollster_trump)
significant_trump_model_2 <- as.data.frame(summary_trump_model_2$coefficients) %>%
  filter(`Pr(>|t|)` < 0.05)

```{r}
#| echo: false
#| eval: true
#| label: tbl-summarystatistics-kamala
#| warning: false
library(modelsummary)
library(knitr)
summary_harris_model_2 <- summary(model_date_pollster)
significant_harris_model_2 <- as.data.frame(summary_harris_model_2$coefficients) %>%
  filter(`Pr(>|t|)` < 0.05)

@fig-prediction-outcome and @fig-prediction-outcome-smooth depict the predicted polling percentages for both candidates over time based on Model 2. The solid lines represent the smoothed predictions for Kamala Harris (blue) and Donald Trump (red). The visualizations reveal that both candidates maintain relatively stable trends, with some fluctuation, particularly influenced by pollsters and specific dates.

@fig-prediction-outcome-smooth indicates that Kamala Harris tends to have a slightly higher polling percentage compared to Donald Trump, with less fluctuation over the observed period. This suggests that her polling results are more consistent across different pollsters and dates.

```{r}
#| echo: false
#| eval: true
#| label: fig-prediction-outcome
#| tbl-cap: "Polling prediction Outcome"
#| warning: false
# Plot model predictions for Trump
# Combine the datasets for Trump and Harris, keeping their respective predictions
combined_data_with_preds <- bind_rows(
  just_harris_high_quality |> mutate(candidate = "Kamala Harris"),
  just_trump_high_quality |> mutate(candidate = "Donald Trump")
)

@@

  # Add labels and theme
  labs(
    x = "Polling End Date ",
    y = "Polling Percentage",
    title = "Polling Predictions for Harris and  Trump",
    color = "Candidate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Plotting the smoothed predictions for both Trump and Harris
```

```{r}
#| echo: false
#| eval: true
#| label: fig-prediction-outcome-smooth
#| tbl-cap: "Polling prediction Outcome"
#| warning: false
ggplot(combined_data_with_preds, aes(x = end_date, y = pct)) +
  # Actual data points with transparency
  geom_point(aes(color = candidate), alpha = 0.3) +

@@

  # Add labels and theme
  labs(
    x = "Polling End Date",
    y = "Polling Percentage",
    title = "Smoothed Polling Predictions for Harris and Trump",
    color = "Candidate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

The smoothed polling predictions in @fig-prediction-outcome-smooth clearly showed that Harris is ahead of Trump throughout the analyzed period. This consistent lead, combined with higher polling percentages reported by major polling organizations such as MassINC Polling Group and University of Maryland/Washington Post, indicates Harris’s current advantage heading toward the election.

@@
